Reinforcement Learning Produces Dominant Strategies for the Iterated Prisoner's Dilemma

### 2017-07-09 main.tex
Add a todo.
Various edits

The Axelrod library \cite{axelrodproject, knight2016open} is an open source
software for conducting IPD research with
reproducibility as a principal goal. Written in the Python programming language,
to date over the library contains source code contributed by over 50 individuals
from a variety of geographic locations and technical backgrounds. The library
is supported by a comprehensive test suite that covers all the intended
behaviors of the strategies in the library, as well as the features that
conduct matches, tournaments, and population dynamics.

@@

In this work we describe how collections of strategies in the Axelrod library
have been used to train new strategies specifically to win IPD tournaments.
These strategies are trained using generic strategy archetypes based on e.g.\
finite state
machines, arriving at particularly effective parameter choices through
evolutionary or particle swarm algorithms. There are several
previous publications that use evolutionary algorithms to
evolve IPD strategies in various circumstances
\cite{ashlock2006training, Ashlock2015, Ashlock2006,
      ashlock2014shaped, Ashlock2014, barlow2015varying,
      fogel1993evolving, marks1989niche, sudo2015effects,
      vassiliades2010multiagent}. See also \cite{Gaudesi2016} for a
strategy trained to win against a collection of well-known IPD opponents and see
\cite{franken2005particle} for a prior use of particle swarm algorithms. Our
results are unique in that we are able to train against a large collection of
well-known strategies available in the scientific literature. Crucially, the
software used in this work is openly available and can be used to train strategies
in the future in a reliable manner, with confidence that the opponent strategies
are correctly implemented, tested and documented.
Moreover, as of the time of writing,
we claim that this work contains the best known strategies for the iterated
prisoner's dilemma.

@@

The Axelrod library now contains many parametrised strategies trained using
machine learning
methods. Most are deterministic, use many rounds of memory, and perform
extremely well in tournaments as will be discussed in Section~\ref{sec:results}.
Training of these strategies will be discussed in Section~\ref{sec:methods}.

These strategies can encode a variety
of other strategies, including classic strategies like Tit For Tat,
handshake strategies, and grudging strategies that always defect after
an opponent defection.

The various archetypes will be described in the following sections.

@@

Training of this strategy corresponds to finding maps from histories to either a
cooperation or a defection.

Although various
combinations of $n_1, m_1,$ and $m_2$ have been tried, the best performance at
the time of
training was obtained for $n_1 = m_1 = m_2 = 2$ and generally for $n_1 > 0$.
A strategy
called EvolvedLookerUp2\_2\_2 is among the top strategies in the library.

@@

Training of this strategy corresponds to finding maps from histories to
a probability of cooperation.

The library includes a strategy trained
with $n_1 = m_1 = m_2 = 2$ that is \emph{mostly deterministic}, with most of the
probabilities being 0 or 1. At one time this strategy outperformed
EvolvedLookerUp2\_2\_2.

@@

\subsection{ANN: Single Layer Artificial Neural Network}

@@

Training of this strategy corresponds to finding parameters of the neural
network.

An inner layer with just five nodes performs quite well in both deterministic and
noisy tournaments. The output of the ANN used in this work is deterministic;
a stochastic variant that outputs probabilities rather than exact moves could
be easily created.

@@

A variant of finite state machine strategies are called hidden Markov models
(HMMs). Like the strategies based on finite state machines, these strategies
also encode an internal state however use probabilistic transitions based on the
prior round of play to other states and cooperate or defect with various
probabilities at each state. This is
shown diagrammatically in Figure~\ref{fig:hmm}.

@@

Training this strategy corresponds to finding mappings of states and histories
to probabilities of cooperating as well as probabilities of the next internal
state.

These are the best performing stochastic strategies
in the library but take longer to train due to their stochasticity.

@@

Last but not least there are several strategies based on ensemble methods that
are common in machine learning called Meta strategies. These strategies are
composed of a team of other strategies. In each round, each member of the team
is polled for its desired next
move. The ensemble then selects the next move based on a rule, such as the
consensus vote in the case of MetaMajority or the best individual performance
in the case of MetaWinner. These strategies were among the best in the library
before the inclusion of those trained by reinforcement learning.

Because these strategies inherit many of the properties of the strategies
on which they are based, including using the match length to defect on the last
rounds of play, not all of these strategies were included in results of this
paper.

@@

All data generated for this work is archived and available...
% TODO Add details once data is archived.

@@

\begin{table}[!hbtp]
        \centering
        \input{./assets/standard_top_15_scores_summary.tex}
        \caption{Standard Tournament: Mean score per turn of top 15 strategies
            (ranked by median over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments).
        The leaderboard is dominated by the trained strategies (indicated by a
        $^{*}$}
        \label{tbl:standard_score}
\end{table}

@@

The tournament results indicate that pre-trained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. An evolutionary algorithm produces strategies based on multiple
standard machine learning techniques that are able to achieve a higher average
score than any other known opponent in a standard tournament. Most of the trained
strategies use multiple rounds of the history of play (some using all of it) and
outperform memory-one strategies (though the trained memory one strategy performs
well). The generic structure of the trained strategies did not appear to be
critical -- strategies based on lookup tables, finite state machines, and stochastic
variants all performed well for standard tournaments.
Single layer neural networks (Section~\ref{sec:ann}) performed well in both
noisy and standard
tournaments
though these had some aspect of human involvement in the selection of features.
This is in line with the other strategies also where human decisions are made
regarding the structure. For the LookerUp and Gambler archetypes
(Sections~\ref{sec:lookerup} and~\ref{sec:gambler})
a decision has to be made
regarding the number of turns of history and initial play that are to be used.
Whilst, the Finite state machines and Hidden Markov models
(Sections\ref{sec:fsm} and~\ref{sec:hmm}) also need input from a
human as to the number of states to use they are much more of a black box and
thus offer wider adaptability.

A lot of strategies could be represented by many archetypes, however some
archetypes will be far more efficient in doing so than
others. The fact that the Lookerup strategy does the best for the standard
tournament indicates that it represents an efficient reduction of
dimension which in turn makes its training more efficient. For the noisy
tournament however the dimension reduction represented by some archetypes
indicates that sufficient features are not captured by the Lookup tables whilst
they are by the
neural network and the finite state machine allowing them to adapt to the noisy
environment.

In opposition to historical tournament results and community folklore,
our results show that complex strategies can be very effective for the
IPD\@. It is not the complexity of strategies that is disadvantageous; rather that directly
designing a broadly effective strategy is no easy task. Of all the human-designed
strategies in the library, only DBS consistently performs well, and it is
substantially more complex than traditional tournament winners like TFT, OmegaTFT,
and zero determinant strategies. Furthermore, dealing with noise is difficult
for most strategies. Two strategies designed specifically to account for noise,
DBS and OmegaTFT, perform well and only DBS performs better than the trained
strategies and \textbf{only} in the noisy context.

Of the strategies trained to maximize their average score all are generally
cooperative, not defecting until the opponent defects. Maximizing for individual
performance across a collection of opponents leads to mutual cooperation despite
the fact that mutual cooperation is an unstable evolutionary equilibrium for the prisoner's
dilemma. Specifically it is noted that the reinforcement learning process for maximizing
payoff does not lead to exploitative zero determinant strategies, which may also
be a result of the collection of training strategies, of which several retaliate harshly.

We take the liberty of generalizing from the results of this study. For the trained
strategies utilizing look up tables we generally found those that incorporate
one or more of the initial rounds of play outperformed those that did not. The
strategies based on neural networks and finite state machines also are able to
condition throughout a match on the first rounds of play. Accordingly, we conclude
that first impressions matter in the IPD\@. The best strategies are nice (never
defecting first) and this property could be further investigated with the
Axelrod library
in future work by e.g.\ forcing all strategies to defect on the first round.

Finally, we note that as the library grows, the top performing strategies
sometimes shuffle, and are not retrained regularly. Most of the strategies were
trained on an earlier version of the library (v2.2.0: \cite{axelrodproject2.2})
that did not include DBS and several other opponents. The precise parameters
that are optimal will depend on the pool of opponents. Moreover we have not
extensively trained strategies to determine the minimum parameters that are
sufficient -- neural networks with fewer nodes and features and finite state
machines with fewer states may suffice. See \cite{ashlock2013impact} for
discussion of resource availability for IPD strategies. It may be possible to
train strategies more effective in noisy tournaments than DBS.

Future work:
% TODO Write the future work section
%* spatial tournaments and other variants
%* Additional strategy archetypes by the Ashlocks, e.g. function stacks, binary
%decision players
%* further refine features and training parameters

>>>>

The Axelrod library \cite{axelrodproject, knight2016open} is an open source
software for conducting IPD research with
reproducibility as a principal goal. Written in the Python programming language,
to date the library contains source code contributed by over 50 individuals
from a variety of geographic locations and technical backgrounds. The library
is supported by a comprehensive test suite that covers all the intended
behaviors of the strategies in the library, as well as the features that
conduct matches, tournaments, and population dynamics.

@@

In this work we describe how collections of strategies in the Axelrod library
have been used to train new strategies specifically to win IPD tournaments.
These strategies are trained using generic strategy archetypes based on e.g.\
finite state
machines, arriving at particularly effective parameter choices through
evolutionary or particle swarm algorithms. There are several
previous publications that use evolutionary algorithms to
evolve IPD strategies in various circumstances
\cite{ashlock2006training, Ashlock2015, Ashlock2006,
      ashlock2014shaped, Ashlock2014, barlow2015varying,
      fogel1993evolving, marks1989niche, sudo2015effects,
      vassiliades2010multiagent}. See also \cite{Gaudesi2016} for a
strategy trained to win against a collection of well-known IPD opponents and see
\cite{franken2005particle} for a prior use of particle swarm algorithms. Our
results are unique in that we are able to train against a large collection of
well-known strategies available in the scientific literature. Crucially, the
software used in this work is openly available and can be used to train strategies
in the future in a reliable manner, with confidence that the opponent strategies
are correctly implemented, tested and documented.
Moreover, as of the time of writing, we claim that this work contains the best
performing strategies for the iterated prisoner's dilemma.

@@

The Axelrod library now contains many parametrised strategies trained using
machine learning
methods. Most are deterministic, use many rounds of memory, and perform
extremely well in tournaments as will be discussed in Section~\ref{sec:results}.
Training of these strategies will be discussed in Section~\ref{sec:methods}.
These strategies can encode a variety
of other strategies, including classic strategies like Tit For Tat,
handshake strategies, and grudging strategies that always defect after
an opponent defection.

@@

Training of this strategy corresponds to finding maps from partial histories to
either a cooperation or a defection. Although various
combinations of $n_1, m_1,$ and $m_2$ have been tried, the best performance at
the time of
training was obtained for $n_1 = m_1 = m_2 = 2$ and generally for $n_1 > 0$.
A strategy
called EvolvedLookerUp2\_2\_2 is among the top strategies in the library.

@@

Training of this strategy corresponds to finding maps from histories to
a probability of cooperation. The library includes a strategy trained
with $n_1 = m_1 = m_2 = 2$ that is \emph{mostly deterministic}, with most of the
probabilities being 0 or 1. At one time this strategy outperformed
EvolvedLookerUp2\_2\_2.

@@

\subsection{ANN: Single Hidden Layer Artificial Neural Network}

@@

Training of this strategy corresponds to finding parameters of the neural
network. An inner layer with just five nodes performs quite well in both deterministic and
noisy tournaments. The output of the ANN used in this work is deterministic;
a stochastic variant that outputs probabilities rather than exact moves could
be easily created.

@@

A variant of finite state machine strategies are called hidden Markov models
(HMMs). Like the strategies based on finite state machines, these strategies
also encode an internal state however use probabilistic transitions based on the
prior round of play to other states and cooperate or defect with various
probabilities at each state. This is
shown diagrammatically in Figure~\ref{fig:hmm}. Training this strategy
corresponds to finding mappings of states and histories to probabilities of
cooperating as well as probabilities of the next internal state.

@@

@@

There are several strategies based on ensemble methods that
are common in machine learning called Meta strategies. These strategies are
composed of a team of other strategies. In each round, each member of the team
is polled for its desired next
move. The ensemble then selects the next move based on a rule, such as the
consensus vote in the case of MetaMajority or the best individual performance
in the case of MetaWinner. These strategies were among the best in the library
before the inclusion of those trained by reinforcement learning. The library
contains strategies containing teams of all the deterministic players, all the
memory-one players, and some others.

Because these strategies inherit many of the properties of the strategies
on which they are based, including using the match length to defect on the last
round(s) of play, not all of these strategies were included in results of this
paper. These strategies do not typically outperform the trained strategies
described above.

@@

All data generated for this work is archived and available...
% TODO Add details once data is archived.
% TODO Include url to repo etc.

@@

\begin{table}[!hbtp]
        \centering
        \input{./assets/standard_top_15_scores_summary.tex}
        \caption{Standard Tournament: Mean score per turn of top 15 strategies
            (ranked by median over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments).
        The leaderboard is dominated by the trained strategies (indicated by a
        $^{*}$)}
        \label{tbl:standard_score}
\end{table}

@@

The tournament results indicate that pre-trained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. An evolutionary algorithm produces strategies based on multiple
generic archetypes that are able to achieve a higher average
score than any other known opponent in a standard tournament. Most of the trained
strategies use multiple rounds of the history of play (some using all of it) and
outperform memory-one strategies from the literature. Interestingly, a trained
memory one strategy produced by a particle swarm algorithm performs well, better
than human designed strategies such as Win Stay Lose Shift and zero determinant
strategies (which enforce a payoff difference rather than maximize total payoff).
The generic structure of the trained strategies did not appear to be
critical for the standard tournament -- strategies based on lookup tables,
finite state machines, neural networks, and stochastic variants all performed well.
Single layer neural networks (Section~\ref{sec:ann}) performed well in both
noisy and standard tournaments
though these had some aspect of human involvement in the selection of features.
This is in line with the other strategies also where some human decisions are made
regarding the structure. For the LookerUp and Gambler archetypes
(Sections~\ref{sec:lookerup} and~\ref{sec:gambler})
a decision has to be made
regarding the number of turns of history and initial play that are to be used.
In contrast, the finite state machines and hidden Markov models
(Sections\ref{sec:fsm} and~\ref{sec:hmm}) required only a choice of the number
of states, and the training algorithm can eliminate unneeded states in the case
of finite state machines (evidenced by the unconnected nodes in the diagrams
for the included representations).

Many strategies can be represented by multiple archetypes, however some
archetypes will be more efficient in encoded the patterns present in the data.
The fact that the Lookerup strategy does the best for the standard
tournament indicates that it represents an efficient reduction of
dimension which in turn makes its training more efficient. In particular the
first rounds of play were valuable bit of information. For the noisy
tournament however the dimension reduction represented by some archetypes
indicates that some features of the data are not captured by the lookup
tables while they are by the neural network and the finite state machine,
allowing them to adapt better to the noisy environment. Intuitively, a noisy
environment can significantly affect a lookup table based on the last two rounds
of play since these action pairs compete with probing defections, apologies, and
retaliations. Accordingly, it is not surprising that additional parameter space
is needed to adapt to a noisy environment.

In opposition to historical tournament results and community folklore,
our results show that complex strategies can be very effective for the
IPD\@. Designing complex strategies for the prisoner's dilemma appears to be
difficult for humans. Of all the human-designed
strategies in the library, only DBS consistently performs well, and it is
substantially more complex than traditional tournament winners like TFT, OmegaTFT,
and zero determinant strategies. Furthermore, dealing with noise is difficult
for most strategies. Two strategies designed specifically to account for noise,
DBS and OmegaTFT, perform well and only DBS performs better than the trained
strategies and \textbf{only} in some noisy contexts. Empirically we find that
DBS (with its default parameters) does not win tournaments at 1% noise.
However DBS has a parameter that accounts for the expected amount of noise and a
followup study with various noise levels could make a more complete study of
the performance of DBS and strategies trained at various noise levels.

The strategies trained to maximize their average score are generally
cooperative and do not defect first. Maximizing for individual
performance across a collection of opponents leads to mutual cooperation despite
the fact that mutual cooperation is an unstable evolutionary equilibrium for the prisoner's
dilemma. Specifically it is noted that the reinforcement learning process for maximizing
payoff does not lead to exploitative zero determinant strategies, which may also
be a result of the collection of training strategies, of which several retaliate
harshly. Training with the objective of maximizing payoff difference may
produce strategies more like zero determinant strategies.

For the trained
strategies utilizing look up tables we generally found those that incorporate
one or more of the initial rounds of play outperformed those that did not. The
strategies based on neural networks and finite state machines also are able to
condition throughout a match on the first rounds of play. Accordingly, we conclude
that first impressions matter in the IPD\@. The best strategies are nice (never
defecting first) and the impact of the first rounds of play could be further
investigated with the Axelrod library
in future work by e.g.\ forcing all strategies to defect on the first round.

Finally, we note that as the library grows, the top performing strategies
sometimes shuffle, and are not retrained automatically. Most of the strategies were
trained on an earlier version of the library (v2.2.0: \cite{axelrodproject2.2})
that did not include DBS and several other opponents. The precise parameters
that are optimal will depend on the pool of opponents. Moreover we have not
extensively trained strategies to determine the minimum parameters that are
sufficient -- neural networks with fewer nodes and features and finite state
machines with fewer states may suffice. See \cite{ashlock2013impact} for
discussion of resource availability for IPD strategies.
